---
title: "P3: Trip Generation"
author: "Jose Ruben Figueroa Velasco"
format: html
editor: visual
---

## Explaining Weekend Home-based Recreational Trips

This is the code that supports the results reported in the memo "Explaining Weekend Home-based Recreational Trips".

## Libraries

The libraries required for the analysis are as follows:

```{r}
#| echo: false
library(tidyverse)
library(here)
library(knitr)
library(srvyr)
library(mlogit)
library(caret)
library(pscl)

here("code",
     "mlogit_helpers.R") |>
  source()
```

## Datasets

The analysis leverages the information available at the household, personal and trip level from the 2017 National Household Travel Survey. We will limit our analysis to the weekends:

```{r}
hh_data <- here("data",
                "NHTS",
                "hhpub.csv") |>
  read_csv(show_col_types = FALSE) |>
  filter(TRAVDAY == "01" | TRAVDAY == "07")

person_data <- here("data",
                    "NHTS",
                    "perpub.csv") |>
  read_csv(show_col_types = FALSE) |>
  filter(TRAVDAY == "01" | TRAVDAY == "07")

trip_data <- here("data",
                    "NHTS",
                    "trippub.csv") |>
  read_csv(show_col_types = FALSE) |>
  filter(TRAVDAY == "01" | TRAVDAY == "07") |>
  select(HOUSEID,
         PERSONID,
         TDTRPNUM,
         TRIPPURP,
         WHYFROM,
         WHYTO,
         TRPTRANS)
```

## Link Up Unlinked Trips

Before filtering for weekend home-based recreational trips, it is important to link up unlinked trips:

```{r}
trip_data <- trip_data |>
  mutate(transfer_trip = WHYTO == "07" | WHYFROM == "07")

transfer_trips <- trip_data |>
  filter(transfer_trip) |>
  filter(WHYTO != WHYFROM) # remove trips that begin AND end with a transfer

next_destination <- c(transfer_trips$WHYTO[2:nrow(transfer_trips)], "last")

transfer_trips <- transfer_trips |>
  mutate(next_destination = next_destination) |>
  mutate(WHYTO = ifelse(WHYTO == "07", next_destination, WHYTO)) |>
  filter(WHYFROM != "07") |>
  select(-next_destination)

trip_data <- trip_data |>
  filter(!transfer_trip) |>
  rbind(transfer_trips) |>
  select(-TRIPPURP, -TRPTRANS, -transfer_trip)
```

## Trip Identification

In this section we identify weekend home-based recreational trips (WHBR) and select the predictors of interest:

```{r}
# Identify weekend home-based recreational trips:
home <- c("01")
recreational <- c("15")

trip_data <- trip_data |>
  mutate(purpose = case_when(WHYTO == home & WHYFROM %in% recreational ~ "WHBR",
                             WHYTO == recreational & WHYFROM %in% home ~ "WHBR",
                             TRUE ~ "OTHER"))

# Predictors:
## Household data (veh_avail, income, rural):
hh_data <- hh_data |>
  mutate(veh_avail = case_when(HHVEHCNT == 0 ~ "Zero",
                               DRVRCNT > HHVEHCNT ~ "Insuff.",
                               TRUE ~ "Suff.")) |>
  mutate(HHFAMINC = as.numeric(HHFAMINC)) |>
  filter(HHFAMINC > 0) |>
  mutate(income = case_when(HHFAMINC < 4 ~ "low",
                             HHFAMINC < 5 & HHSIZE > 1 ~ "low",
                             HHFAMINC < 6 & HHSIZE > 3 ~ "low",
                             HHFAMINC < 7 & HHSIZE > 5 ~ "low",
                             HHFAMINC < 8 & HHSIZE > 7 ~ "low",
                             HHFAMINC > 8 ~ "high",
                            TRUE ~ "middle")) |>
  mutate(income = factor(income, levels = c("low", "middle", "high"))) |>
  mutate(urban = case_when(URBRUR == "01" ~ "urban",
                           TRUE ~ "rural")) |>
  select(HOUSEID,
         veh_avail,
         income,
         urban)

## Personal data (sex, race, age)
persons <- person_data |>
  select(HOUSEID, PERSONID, R_SEX, R_RACE, R_AGE) |>
  mutate(male = R_SEX == "01") |>
  mutate(white = R_RACE == "01") |>
  inner_join(hh_data)

# Count WHBR trips and merge datasets:
whbr_trips <- trip_data |>
  filter(purpose == "WHBR") |>
  group_by(HOUSEID, PERSONID) |>
  summarise(whbr_trips = n())

persons <- persons |>
  left_join(whbr_trips) |>
  replace_na(list(whbr_trips = 0)) |>
  mutate(income = factor(income, levels = c("low", "middle", "high")))
```

## Sample Visualization

The distribution of the number of WHBR trips is as follows:

```{r}
ggplot(persons) +
  geom_histogram(aes(x = whbr_trips),
                 binwidth = 1,
                 color = "gray",
                 fill = "lightblue") +
  scale_x_continuous(name = "Number of weekend home-based recreational trips",
                     breaks = seq(0, 12, by=1)) +
  scale_y_continuous(name = "Number of individuals in sample") +
  theme_minimal()
```

Given the distribution of the number of WHBR trips, either a Poisson regression or a negative binomial regression appears to be more appropriate than a simple linear regression to explore the data. To explore which regression is required, we will perform the following test:

```{r}
persons |>
  summarise(mean_whbr_trips = mean(whbr_trips),
            st_dev_whbr_trips = sd(whbr_trips)) |>
  kable(digits = 3)
```

Since the standard deviation is much greater than the mean of the number of WHBR trips, we will use a negative binomial regression to perform the analysis.

## Model Estimation

```{r}
library(MASS)

persons$R_AGE2 <- persons$R_AGE^2
whbr_model <- glm.nb(whbr_trips ~ 
                     veh_avail + 
                     income + 
                     urban +
                     male + 
                     white +
                     R_AGE + 
                     R_AGE2, 
   data = persons)

summary(whbr_model)
```

The theta parameter is close to zero, suggesting that the number of WHBR trips presents over-dispersion and, thus, the negative binomial model is appropriate.
